The Calculus of Linear Constructions

### 2021-10-15 writing/draft3.tex
draft

% document body
\begin{document}
  \maketitle 
  \begin{abstract}
    The Calculus of Linear Constructions (CLC) is an extension of the Calculus of Constructions (CC) with linear types. Specifically, CLC extends CC with a hierarchy of linear universes, and an indexed typing judgment that precisely controls the weakening and contraction of its term level inhabitants. We study the meta-theory of CLC, showing that it is a sound logical framework for reasoning about resource. CLC is backwards compatible with CC, allowing CLC to enjoy the fruits of decades of CC research. We have formalized and proven correct all major results of the core calculus in the Coq Proof Assistant. We extend CLC with linear inductive types and show that CLC as a programming language enables the manipulation of mutable data structures in a principled way.
  \end{abstract}

  \section{Introduction}
  The Calculus of Constructions (CC) is a dependent type theory introduced by Coquand, and Huet in their landmark work \cite{cc}. In CC types can depend on terms, allowing one to write precise specifications as types. Today, CC and its variations CIC \cite{cic} and ECC \cite{ecc} lie at the core of popular proof assistants such as Coq \cite{coq}, Agda \cite{agda}, Lean \cite{lean}, and others. These theorem provers have found great success in the fields of software verification \cite{compcert,deepspec}, and constructive mathematics \cite{four-color,schemes}.

  However, due to its origins as a logical framework for constructive mathematics, it is quite difficult for CC to encode and reason about resources. Intuitively, a mathematical theorem can be applied an unrestricted number of times. Comparatively, the usage of resources is more limited. For example, if we encode Girard's classical example \cite{girard95} of purchasing cigarettes literally into CC as a function of type:
  \begin{equation*}
    Money \rightarrow Camels + Marlboro
  \end{equation*}
  If viewed as a propositional implication, the customer will still maintain full ownership of their money after paying the vendor, because implication does not diminish the validity of its antecedent. Unless the vendor is exceedingly generous, we are faced with the crime of counterfeiting. Users of proof assistants based on CC often need to embed external logics \cite{iris} to provide additional reasoning principles for dealing with resource. The design and embedding of these logics are a difficult problem in their own right, requiring additional proofs to justify its soundness. We propose an alternative solution: extend CC with linear types.

@@

  Work have been done to extend dependent type theories with linear types. Cervesato and Pfenning extends the Edinburgh Logical Framework with linear types \cite{lf,llf}, being the first to demonstrate that dependent types and linear types can coexist within a type theory. V\'{a}k\'{a}r \cite{vakar14} gives a categorical semantics for linear dependent types.  Krishnaswami et al. present a dependent linear type theory \cite{neel15} based on Benton's early work of mixed linear and non-linear calculus, demonstrating the ability to internalize imperative programming the style of Hoare Type Theory \cite{htt}. Luo et al. introduce the property of essential linearity, and a mixed linear/non-linear context, describing the first type theory that allows types to depend on linear terms. Based on initial ideas of McBride \cite{nothing}, Atkey's Quantitative Type Theory (QTT) \cite{qtt} uses semi-ring annotations to track variable occurrence, simulating irrelevance, linear, and affine types within a unified framework. The Idris 2 programming language \cite{idris2} implements QTT as its core type system.

  We propose a new linear dependent type system - The Calculus of Linear Constructions (CLC). CLC extends CC$\omega$ with linear types. CC$\omega$ itself is an extension of CC with a cumulative hierarchy of type universes. We add extra universes $L$ of linear types with cumulativity parallel to the universes $U$ of non-linear types. Universe information is propagated by an indexed typing judgment down to the term level, controlling the usage of weakening and contraction rules. This ultimately results in the \textit{linearity} theorem, stating that all resources are used exactly once.

  The presence of both linear and dependent types enables CLC to write specifications that faithfully encodes the usage of resource. The previous example of monetary transaction can be refined using an indexed linear type family $Money : \mathbb{N} \rightarrow L$ as follows.
  \begin{equation*}
    (Money\ 5) \arw{L}{L} (Camels + Marlboro)
  \end{equation*}
  This new specification for transaction states that it requires a payment of 5 units of money, the customer is relieved of their ownership after the transaction finishes, effectively preventing the contradiction of having your cake and eating it too.

  Compared to preexisting approaches for integrating linear types and dependent types, CLC offers a ``lightweight'' approach to extending CC$\omega$ with linear types, akin to Mazurak et al.'s work on System F \cite{mazurak}. This allows for a straightforward modeling of CLC in CC$\omega$, and lifting of CC$\omega$ into CLC, endowing CLC with the fruits of decades of CC research. We further extend CLC with inductive types, showing that as a programming language it can manipulate mutable data structures in a principled way. We have formalized all major results in Coq, and implemented a prototype in OCaml.

@@

  \noindent \textbf{\textit{Contributions}}: 
  Our contributions can be summarized as follows.
  \begin{itemize}
    \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
    \item Next, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, propagation, and subject reduction. 
    \item We observe that CLC is highly backwards compatible with CC$\omega$. We construct a reduction preserving model of CLC in CC$\omega$, showing that CLC is consistent. 
    \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
    \item Furthermore, we extend CLC with linear inductive data, demonstrating that as a programming language, CLC can safely manipulate mutable data structures.
    \item Finally, we give an implementation extended with user definable linear and non-linear inductive types. Algorithmic type checking employed by the implementation streamlines the process of writing CLC.
  \end{itemize}

@@

      <$m, n, A, B, C$> ::= $U_i$ | $L_k$ | $x$ \hspace*{7.3em} expressions
      \indalt $(x : A) \arw{s}{t} B$
      \indalt $(x : A) \lrw{s}{t} B$
      \indalt $\lambda x. n$ | $m\ n$
    \end{grammar}
    \end{minipage}
    \label{syntax}
  \end{figure}
  A clear departure of our language from standard presentations of both linear type theory, and dependent type theory are the indexed function types: $(x : A) \arw{s}{t} B$, $(x : A) \lrw{s}{t} B$. The reason for these indices is that we have built the ! exponential of linear logic directly into universe sorts. The indices $s$, and $t$ annotate the domain, and co-domain's respective universe sort, hence annotating linearity. The behavior of ! is difficult to account for even in simple linear type theory. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into universe sorts, we implicitly limit ! to only be used canonically. This allows us to derive the substitution lemma, and construct a direct modeling of CLC in CC$\omega$ without needing any machinery for manipulating exponential.
  \begin{figure}[H]
    \caption{Correspondence of CLC types and MELL implications}
    \begin{align}
      (\_ : A) \arw{U}{U} B \quad &\Leftrightarrow \quad !(!A \multimap\ !B) \\
      (\_ : A) \arw{U}{L} B \quad &\Leftrightarrow \quad !(!A \multimap B) \\
      (\_ : A) \arw{L}{U} B \quad &\Leftrightarrow \quad !(A \multimap\ !B) \\
      (\_ : A) \arw{L}{L} B \quad &\Leftrightarrow \quad !(A \multimap B) \\
      (\_ : A) \lrw{U}{U} B \quad &\Leftrightarrow \quad !A \multimap\ !B \\
      (\_ : A) \lrw{U}{L} B \quad &\Leftrightarrow \quad !A \multimap B \\
      (\_ : A) \lrw{L}{U} B \quad &\Leftrightarrow \quad A \multimap\ !B \\
      (\_ : A) \lrw{L}{L} B \quad &\Leftrightarrow \quad A \multimap B
    \end{align}
    \label{correspondence}
  \end{figure}
  Figure \ref{correspondence} illustrates the correspondence between CLC function types and Multiplicative Exponential Linear Logic (MELL) implications. MELL lacks counterparts for the cases (1), (2), (5), (6) if the co-domain $B$ is dependent on arguments of domain $A$. We will discuss function type formation in Section \ref{tyformation} in greater detail.

  In practice, algorithmic type checking techniques allow users to omit writing most of these sort indices. Our implementation employs bi-directional type checking and users never interact with indices in the surface syntax.

@@

  \begin{definition}
    The cumulativity relation ($\preceq$) is the smallest binary relation over terms such that
    \begin{enumerate}
      \item $\preceq$ is a partial order with respect to definitional equality.
        \begin{enumerate}
          \item If $A \equiv B$, then $A \preceq B$.
          \item If $A \preceq B$ and $B \preceq A$, then $A \equiv B$.
          \item If $A \preceq B$ and $B \preceq C$, then $A \preceq B$.
        \end{enumerate}
      \item $U_* \preceq U_0 \preceq U_1 \preceq U_2 \preceq \cdots$
      \item $L_0 \preceq L_1 \preceq L_2 \preceq \cdots$
      \item If $A_1 \equiv A_2$ and $B_1 \preceq B_2$, then
        $(x : A_1) \arw{s}{t} B_1 \preceq (x : A_2) \arw{s}{t} B_2$
      \item If $A_1 \equiv A_2$ and $B_1 \preceq B_2$, then
        $(x : A_1) \lrw{s}{t} B_1 \preceq (x : A_2) \arw{s}{t} B_2$
    \end{enumerate}
  \end{definition}

@@

  \subsection{Typing Judgment}
  Typing judgments in CLC take on the form of $\Gamma \vdash m \stype{s} A$. Intuitively, this judgment states that the term $m$ is an inhabitant of type $A$, with free variables typed in $\Gamma$. The sort index $s$ tells us the linearity of $m$. Specifically, if $s = U$, then $m$ has unrestricted usage. Conversely, if $s = L$, then $m$ must be used exactly once. Theorem \ref{propagation} (propagation) shows that $s$ corresponds exactly to the sort of $A$'s universe.

  \begin{definition} We formally define the terms \textit{non-linear}, \textit{linear}, \textit{unrestricted}, and \textit{restricted}.
    \begin{enumerate}
      \item A \textit{type} $A$ is \textit{non-linear} under context $\Gamma$ if $\Gamma \vdash A \utype U_i$.
      \item A \textit{type} $A$ is \textit{linear} under context $\Gamma$ if $\Gamma \vdash A \utype L_k$.
      \item A \textit{term} $m$ is \textit{unrestricted} under context $\Gamma$ if it has type $A$, and $\Gamma \vdash m \utype A$. A unrestricted term may be used an arbitrary number of times.
      \item A \textit{term} $m$ is \textit{restricted} under context $\Gamma$ if it has type $A$, and $\Gamma \vdash m \ltype A$ A restricted term must be used exactly once.
    \end{enumerate}
  \end{definition}

>>>>

% document body
\begin{document}
  \maketitle 
  \begin{abstract}
    The Calculus of Linear Constructions (CLC) is an extension of the Calculus of Constructions (CC) with linear types. Specifically, CLC extends CC with a hierarchy of linear universes that precisely controls the weakening and contraction of its term level inhabitants. We study the meta-theory of CLC, showing that it is a sound logical framework for reasoning about resource. CLC is backwards compatible with CC, allowing CLC to enjoy the fruits of decades of CC research. We have formalized and proven correct all major results of the core calculus in the Coq Proof Assistant. We extend CLC with linear inductive types, showing that CLC facilitates correct by construction imperative programming.
  \end{abstract}

  \section{Introduction}
  The Calculus of Constructions (CC) is a dependent type theory introduced by Coquand, and Huet in their landmark work \cite{cc}. In CC types can depend on terms, allowing one to write precise specifications as types. Today, CC and its variations CIC \cite{cic} and ECC \cite{ecc} lie at the core of popular proof assistants such as Coq \cite{coq}, Agda \cite{agda}, Lean \cite{lean}, and others. These theorem provers have found great success in the fields of software verification, and constructive mathematics.

  However, due to its origins as a logical framework for constructive mathematics, it is quite difficult for CC to encode and reason about resources. Intuitively, a mathematical theorem can be applied an unrestricted number of times. Comparatively, the usage of resources is more limited. For example, if we encode Girard's classical example \cite{girard95} of purchasing cigarettes literally into CC as a function of type:
  \begin{equation*}
    Money \rightarrow Camels + Marlboro
  \end{equation*}
  If viewed as a propositional implication, the customer will still maintain full ownership of their money after paying the vendor, because implication does not diminish the validity of its antecedent. Unless the vendor is exceedingly generous, we are faced with the crime of counterfeiting. Users of proof assistants based on CC often need to embed external logics such as Separation Logic to provide additional reasoning principles for dealing with resource. The design and embedding of these logics are a difficult problem in their own right, requiring additional proofs to justify their soundness, and effectiveness. We propose an alternative solution: extend CC with linear types.

@@

  Work has been done to extend dependent type theories with linear types. Cervesato and Pfenning extend the Edinburgh Logical Framework with linear types \cite{lf,llf}, being the first to demonstrate that dependent types and linear types can coexist within a type theory. V\'{a}k\'{a}r \cite{vakar14} presents a linear dependent type theory, with syntax and semantics drawing inspiration from DILL \cite{dill}.  Krishnaswami et al. present a dependent linear type theory \cite{neel15} based on Benton's earlier work on mixed linear and non-linear calculus, demonstrating the ability to internalize imperative programming in the style of Hoare Type Theory \cite{htt}. Luo et al. introduce the property of essential linearity, and a mixed linear/non-linear context, describing the first type theory that allows types to depend on linear terms. Based on initial ideas of McBride \cite{nothing}, Atkey's Quantitative Type Theory (QTT) \cite{qtt} uses semi-ring annotations to track variable occurrence, simulating irrelevance, linear, and affine types within a unified framework. The Idris 2 programming language \cite{idris2} implements QTT as its core type system.

  We propose a new linear dependent type system - The Calculus of Linear Constructions (CLC). CLC extends CC$\omega$ with linear types. CC$\omega$ itself is an extension of CC with a cumulative hierarchy of type universes. We add extra universes $L$ of linear types with cumulativity parallel to the universes $U$ of non-linear types. This ultimately cumulates in the \textit{linearity} theorem, stating that all resources are used exactly once.

  The presence of both linear and dependent types enables CLC to write specifications that faithfully encode the usage of resources. The previous example of monetary transaction can be refined using an indexed linear type family $Money : \mathbb{N} \rightarrow L$ as follows.
  \begin{equation*}
    (x \ltype Money\ 5) \rightarrow (Camels + Marlboro)
  \end{equation*}
  This new specification for transaction states that it requires a payment of 5 units of money, the customer is relieved of their ownership after the transaction finishes, effectively preventing the contradiction of having your cake and eating it too.

  Compared to preexisting approaches for integrating linear types and dependent types, CLC offers a ``lightweight'' approach to extending CC$\omega$ with linear types, akin to Mazurak et al.'s work on System F \cite{mazurak}. This allows for a straightforward modeling of CLC in CC$\omega$, and lifting of CC$\omega$ into CLC, endowing CLC with the fruits of decades of CC research. We further extend CLC with inductive types, showing that CLC facilitates correct by construction imperative programming. We have formalized all major results in Coq, and implemented a prototype in OCaml.

@@

  \noindent \textbf{\textit{Contributions}}: 
  Our contributions can be summarized as follows.
  \begin{itemize}
    \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
    \item Next, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, propagation, and subject reduction. 
    \item We observe that CLC is highly backwards compatible with CC$\omega$, enabling us to construct a reduction preserving model of CLC in CC$\omega$. This shows that CLC is consistent. 
    \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
    \item Furthermore, we extend CLC with linear inductive data, demonstrating CLC capabilities for correct by construction imperative programming.
    \item Finally, we give an implementation extended with user definable linear and non-linear inductive types. Algorithmic type checking employed by the implementation streamlines the process of writing CLC.
  \end{itemize}

@@

      <$m, n, A, B, C$> ::= $U_i$ | $L_k$ | $x$ \hspace*{7.3em} expressions
      \indalt $(x :_s A) \rightarrow B$
      \indalt $(x :_s A) \multimap B$
      \indalt $\lambda x. n$ | $m\ n$
    \end{grammar}
    \end{minipage}
    \label{syntax}
  \end{figure}
  A clear departure of our language from standard presentations of both linear type theory, and dependent type theory are the presence of two function types: $(x :_s A) \rightarrow B$, $(x :_s A) \multimap B$. The reason for these function types is that we have built the ! exponential of linear logic directly into universe sorts. The sort annotation $s$ here records the universe sort of the function domain. The behavior of ! is difficult to account for even in simple linear type theory. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into universe sorts, we implicitly limit ! to only be canonically usable. This allows us to derive the substitution lemma, and construct a direct modeling of CLC in CC$\omega$ without needing any machinery for manipulating exponential.
  \begin{figure}[H]
    \caption{Correspondence of CLC types and MELL implications}
    \begin{align}
      (\_ \utype A) \rightarrow B \quad &\Leftrightarrow \quad !(!A \multimap B) \\
      (\_ \ltype A) \rightarrow B \quad &\Leftrightarrow \quad !(A \multimap B) \\
      (\_ \utype A) \multimap B \quad &\Leftrightarrow \quad !A \multimap B \\
      (\_ \ltype A) \multimap B \quad &\Leftrightarrow \quad A \multimap B
    \end{align}
    \label{correspondence}
  \end{figure}
  Figure \ref{correspondence} illustrates the correspondence between CLC function types and Multiplicative Exponential Linear Logic (MELL) implications. MELL lacks counterparts for the cases (1), (3) if the co-domain $B$ is dependent on arguments of domain $A$. We will discuss function type formation in Section \ref{tyformation} in greater detail.

  In practice, algorithmic type checking techniques allow users to omit writing sort indices. Our implementation employs bi-directional type checking and users never interact with indices in the surface syntax.

@@

  \begin{definition}
    The cumulativity relation ($\preceq$) is the smallest binary relation over terms such that
    \begin{enumerate}
      \item $\preceq$ is a partial order with respect to definitional equality.
        \begin{enumerate}
          \item If $A \equiv B$, then $A \preceq B$.
          \item If $A \preceq B$ and $B \preceq A$, then $A \equiv B$.
          \item If $A \preceq B$ and $B \preceq C$, then $A \preceq B$.
        \end{enumerate}
      \item $U_* \preceq U_0 \preceq U_1 \preceq U_2 \preceq \cdots$
      \item $L_0 \preceq L_1 \preceq L_2 \preceq \cdots$
      \item If $A_1 \equiv A_2$ and $B_1 \preceq B_2$, then
        $(x \stype{s} A_1) \rightarrow B_1 \preceq (x \stype{s} A_2) \rightarrow B_2$
      \item If $A_1 \equiv A_2$ and $B_1 \preceq B_2$, then
        $(x \stype{s} A_1) \multimap B_1 \preceq (x \stype{s} A_2) \multimap B_2$
    \end{enumerate}
  \end{definition}

@@

  \subsection{Typing Judgment}
  Typing judgments in CLC take on the form of $\Gamma \vdash m : A$. Intuitively, this judgment states that the term $m$ is an inhabitant of type $A$, with free variables typed in $\Gamma$.

  \begin{definition} We formally define the terms \textit{non-linear}, \textit{linear}, \textit{unrestricted}, and \textit{restricted}.
    \begin{enumerate}
      \item A \textit{type} $A$ is \textit{non-linear} under context $\Gamma$ if $\Gamma \vdash A : U_i$.
      \item A \textit{type} $A$ is \textit{linear} under context $\Gamma$ if $\Gamma \vdash A : L_k$.
      \item A \textit{term} $m$ is \textit{unrestricted} under context $\Gamma$ if there exists non-linear type $A$, such that $\Gamma \vdash m : A$. A unrestricted term may be used an arbitrary number of times.
      \item A \textit{term} $m$ is \textit{restricted} under context $\Gamma$ if there exists linear type $A$, such that $\Gamma \vdash m : A$. A restricted term must be used exactly once.
    \end{enumerate}
  \end{definition}
